rang: Reconstructing reproducible R computational environments

### 2023-03-10 paper/paper.qmd
Fix formatting issues #111

> ~ excerpt of [this README file](https://www.michelecoscia.com/?page_id=2105)

Other than bad programming practices [@trisovic:2022], the main computing barrier to computational reproducibility is the failure to reconstruct the computational environment like the one used by the original researchers. This task looks trivially simple. But as computer science research has shown, this task is incredibly complex [@abate:2015:MCR;@dolstra:2010:N]. In the realm of a usual scripting language such as R [^r], that pertains four aspects: a) operating system, b) system components such as `libxml2`, c) the exact R version, and d) what and which version of the installed R packages. We will call them Component A, B, C, D in the following sections. Any change in these four components can possibly affect the execution of any shared computer code. For example, the lack of the system component `libxml2` can impact whether the R package `xml2` can be installed on a Linux system. If the shared computer code requires the R package `xml2`, then the whole execution fails.

@@

There has been many papers written on how containerization solutions such as Docker can be helpful also to foster computational reproducibility of science [e.g. @nuest:2019;@peikert:2021:RDA;@boettiger:2017:IR]. Although tutorials are available [e.g. @nuest:2019], providing a declarative description of the computational environment in the form of Dockerfile is far from the standard code sharing practice. This might be due to a lack of (DevOps) skills of most scientists to create a Dockerfile [@kim:2018:E]. But there are many tools available to automate the process [e.g. @nuest:2019]. The case in point described in this paper, `rang`, is one of them. We argue that `rang` is the only easy-to-use solution available that can pin down and restore all four components without the reliance on any commercial service such as MRAN.

@@

It is also worth mentioning that MRAN is not the only archival service. Posit also provides a free (*gratis*) time-stamped daily backup of CRAN and Bioconductor (a series of repositories of R package for bioinformatics and computational biology) called Posit Public Package Manager (https://packagemanager.rstudio.com/client/#/repos/2/packages/). It can goes as far back as October 10, 2017.

@@

## Structure of this paper

In Section 2, we will explain how to use `rang`. In Section 3, `rang` is used to enable the reproducibility of published literature (with increasing sophistication). However, one can still use `rang` for prospective usage and arguably can ensure a longer term computational reproducibility than other solutions. In Section 4, `rang` is used to create an executable research compendium with Docker and `Make` [@baker:2020:UGM].

@@

[^1]: https://stat.ethz.ch/pipermail/r-announce/1997/000001.html

@@

[^pkgrefs]: https://r-lib.github.io/pkgdepends/reference/pkg_refs.html

@@

[^hilgard]: https://github.com/Joe-Hilgard/vvg-2d4d

@@

[^groundhog]: http://datacolada.org/100

@@

[^checkerror]: https://cran-archive.r-project.org/web/checks/2019/2019-03-05_check_results_maxent.html

[^evidence]: https://github.com/cran/maxent/commit/9d46c6aad27a1f41a78907b170ddd9a586192be9

@@

[^taskview]: https://cran.r-project.org/web/views/SpatioTemporal.html

Even with this over-a-decade removal and new packages with similar functionalities have been created, there is evidence that `ptproc` is still being sought for. As late as 2017, there are blog posts on how to install the long obsolete package on modern versions of R [^blog]. The package is extremely challenging to install on a modern R system because the package was written before the introduction of name space management in R 1.7.0 [@RN-2003-001]. In other words, the available tarball file from the original author's website does not contain a `NAMESPACE` file as all other modern R packages do.

@@

[^peng]: https://www.biostat.jhsph.edu/~rpeng/software/

[^blog]: https://blog.mathandpencil.com/installing-ptproc-on-osx and https://tomaxent.com/2017/03/16/Installing-ptproc-on-Ubuntu-16-04-LTS/

@@

[^sushi]: https://github.com/PhanstielLab/Sushi/blob/master/vignettes/PaperFigure.R

@@

The above R script is saved as `oser.R`. The central piece of the executable compendium is the `Makefile`.

```Makefile
output_file=reproduced.html
r_cmd = "rmarkdown::render('materials/README.Rmd', output_file = '${output_file}')"
handle=oser
local_file=${handle}_README.html

@@

[^zenodo]: The complete version of the executable compendium is available from Zenodo: https://doi.org/10.5281/zenodo.7708417

@@

[^singularity]: https://docs.sylabs.io/guides/3.0/user-guide/singularity_and_docker.html

@@

Computational reproducibility is a complex topic and as in all of these complex topics, there is no silver bullet [@canon:2019:CPR]. All solutions have their trade-offs. The (re)construction process based on `rang` takes notably more time than other solutions because all packages are compiled from source. `rang` trades computational efficiency of this often one-off (re)constructing process for correctness, backward compatibility and independence from any commercial backups of software repositories such as MRAN. There are also other limitations. In the Vignette of `rang` (https://cran.r-project.org/web/packages/rang/vignettes/faq.html), we list all of these limitations as well as possible mitigation.

>>>>

> ~ excerpt of [the README file](https://www.michelecoscia.com/?page_id=2105) of a software for polarization estimation

Other than bad programming practices [@trisovic:2022], the main computing barrier to computational reproducibility is the failure to reconstruct the computational environment like the one used by the original researchers. This task looks trivially simple. But as computer science research has shown, this task is incredibly complex [@abate:2015:MCR;@dolstra:2010:N]. For a usual scripting language such as R [^r], that pertains to four aspects: A) operating system, B) system components such as `libxml2`, C) the exact R version, and D) what and which version of the installed R packages. We will call them Component A, B, C, D in the following sections. Any change in these four components can possibly affect the execution of any shared computer code. For example, the lack of the system component `libxml2` can impact whether the R package `xml2` can be installed on a Linux system. If the shared computer code requires the R package `xml2`, then the whole execution fails.

@@

There has been many papers written on how containerization solutions such as Docker can be helpful also to foster computational reproducibility of science [e.g. @nuest:2019;@peikert:2021:RDA;@boettiger:2017:IR]. Although tutorials are available [e.g. @nuest:2019], providing a declarative description of the computational environment in the form of Dockerfile is far from the standard code sharing practice. This might be due to a lack of (DevOps) skills of most scientists to create a Dockerfile [@kim:2018:E]. But there are many tools available to automate the process [e.g. @nuest:2019]. The case in point described in this paper, `rang`, is one of them. We argue that `rang` is the only easy-to-use solution available that can pin down and restore all four components without the reliance on any commercial service <!-- such as MRAN -->.

@@

It is also worth mentioning that MRAN is not the only archival service. Posit also provides a free (*gratis*) time-stamped daily backup of CRAN and Bioconductor (a series of repositories of R package for bioinformatics and computational biology) called Posit Public Package Manager [^posit]. It can goes as far back as October 10, 2017.

[^posit]: [https://packagemanager.rstudio.com/client/#/repos/2/packages/](https://packagemanager.rstudio.com/client/#/repos/2/packages/)

@@

<!-- ## Structure of this paper -->

<!-- In Section 2, we will explain how to use `rang`. In Section 3, `rang` is used to enable the reproducibility of published literature (with increasing sophistication). However, one can still use `rang` for prospective usage and arguably can ensure a longer term computational reproducibility than other solutions. In Section 4, `rang` is used to create an executable research compendium with Docker and `Make` [@baker:2020:UGM]. -->

@@

[^1]: [https://stat.ethz.ch/pipermail/r-announce/1997/000001.html](https://stat.ethz.ch/pipermail/r-announce/1997/000001.html)

@@

[^pkgrefs]: [https://r-lib.github.io/pkgdepends/reference/pkg_refs.html](https://r-lib.github.io/pkgdepends/reference/pkg_refs.html)

@@

[^hilgard]: [https://github.com/Joe-Hilgard/vvg-2d4d](https://github.com/Joe-Hilgard/vvg-2d4d)

@@

[^groundhog]: [http://datacolada.org/100](http://datacolada.org/100)

@@

[^checkerror]: [https://cran-archive.r-project.org/web/checks/2019/2019-03-05_check_results_maxent.html](https://cran-archive.r-project.org/web/checks/2019/2019-03-05_check_results_maxent.html)

[^evidence]: [https://github.com/cran/maxent/commit/9d46c6aad27a1f41a78907b170ddd9a586192be9](https://github.com/cran/maxent/commit/9d46c6aad27a1f41a78907b170ddd9a586192be9)

@@

[^taskview]: [https://cran.r-project.org/web/views/SpatioTemporal.html](https://cran.r-project.org/web/views/SpatioTemporal.html)

Even with this over-a-decade removal and new packages with similar functionalities have been created, there is evidence that `ptproc` is still being sought for. As late as 2017, there are blog posts on how to install the long obsolete package on modern versions of R [^blog]. The package is extremely challenging to install on a modern R system because the package was written before the introduction of name space management in R 1.7.0 [@RN-2003-001]. In other words, the available tarball files from the original author's website and CRAN do not contain a `NAMESPACE` file like all other modern R packages do.

@@

[^peng]: [https://www.biostat.jhsph.edu/~rpeng/software/](https://www.biostat.jhsph.edu/~rpeng/software/)

[^blog]: [https://blog.mathandpencil.com/installing-ptproc-on-osx](https://blog.mathandpencil.com/installing-ptproc-on-osx) and [https://tomaxent.com/2017/03/16/Installing-ptproc-on-Ubuntu-16-04-LTS/](https://tomaxent.com/2017/03/16/Installing-ptproc-on-Ubuntu-16-04-LTS/)

@@

[^sushi]: [https://github.com/PhanstielLab/Sushi/blob/master/vignettes/PaperFigure.R](https://github.com/PhanstielLab/Sushi/blob/master/vignettes/PaperFigure.R)

@@

The above R script is saved as `oser.R`. The central piece of the executable compendium is the `Makefile` [@baker:2020:UGM].

```Makefile
output_file=reproduced.html
r_cmd = "rmarkdown::render('materials/README.Rmd',\
output_file = '${output_file}')"
handle=oser
local_file=${handle}_README.html

@@

[^zenodo]: The complete version of the executable compendium is available on Zenodo: [https://doi.org/10.5281/zenodo.7708417](https://doi.org/10.5281/zenodo.7708417)

@@

[^singularity]: [https://docs.sylabs.io/guides/3.0/user-guide/singularity_and_docker.html](https://docs.sylabs.io/guides/3.0/user-guide/singularity_and_docker.html)

@@

Computational reproducibility is a complex topic and as in all of these complex topics, there is no silver bullet [@canon:2019:CPR]. All solutions have their trade-offs. The (re)construction process based on `rang` takes notably more time than other solutions because all packages are compiled from source. `rang` trades computational efficiency of this often one-off (re)constructing process for correctness, backward compatibility and independence from any commercial backups of software repositories such as MRAN. There are also other limitations. In the Vignette of `rang` ([https://cran.r-project.org/web/packages/rang/vignettes/faq.html](https://cran.r-project.org/web/packages/rang/vignettes/faq.html)), we list all of these limitations as well as possible mitigation.
